### 2014-10-16 paper/include/background.tex

The DRAM cell array in each bank is structured as in \reffig{fig:dram_orgb}. One cell consists of a transistor and a capacitor, where the transistor is controlled by the \textit{word line} wire and connects the capacitor to the \textit{bit line} wire. The word line connects a number of cells which forms a row. Each bit line is connected to a sense amplifier, and the row of sense amplifiers constitutes the \textit{row buffer}. Data is stored as a charge on the cell capacitor.

@@

To read a row, the word line connected to the row's transistors must first be activated. When the word line is activated, the transistors connect the capacitors to the bit lines. When the capacitors connect to the bit lines, charge will either flow from or to the capacitors depending on whether they were charged or not, respectively. For this to work, the bit lines have to be precharged to $V_{DD}/2$. The sense amplifiers then detect the change of voltage, which makes it possible to interpret the stored data. The charge stored on the capacitors has, as mentioned, left the capacitor and a read operation is thus destructive. To maintain the data the capacitors are recharged by the bit lines, which are driven to either either $V_{DD}/2$ or $0\:V$ by the sense amplifiers depending on the former voltage detected. Finally, when another row on the bank is accessed, the open row has to be closed, which disconnects the capacitors from the bit lines.

To write to a row, the corresponding word line is first activated. As during a read, the transistors then connect the capacitors to the bit lines, and charge will either flow from or to the capacitor. The row buffer is then loaded with the data that are driven to the bit lines by the sense amplifiers, thus storing the information on the capacitors. The row will be closed when another row is accessed.

@@

To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the $64\:ms$ interval. This approach is called \textit{burst refresh}. The approach blocks normal accesses for a long period, and power consumption peaks during the burst time. A better alternative is to use \textit{distributed refresh}, which scatter the refreshes across the interval, resulting in a lower average latency for the processor and more stable power consumption.

The implementation of a DRAM refresh cycle can be made in two different ways, to either affect one row or many. The first is called \textit{RAS-only refresh} (ROR) where a refresh cycle refreshes one targeted row. To target one row, the row's address is put on the bus, and the MC have the responsibility that a row is refreshed within the maximum refresh period. The second is called \textit{CAS before RAS refresh} (CBR). In this scheme, the memory module have an internal address counter that displays which row that will be refreshed. On a refresh cycle, the row stated by the internal address counter is refreshed and the counter is then incremented. Contrary to ROR, CBR do not need provided addresses on the bus, which results in a lower power consumption and less bus usage. % During CBR, a row in each bank is updated, right? if so. write that in the text.

>>>>

The DRAM cell array in each bank is structured as in \reffig{fig:dram_orgb}. One cell consists of a transistor and a capacitor, where the transistor is controlled by the \textit{word line} wire and connects the capacitor to the \textit{bit line} wire. The word line connects a number of cells which forms a row. Each bit line is connected to a sense amplifier, and the row of sense amplifiers constitute the \textit{row buffer}. Data is stored as a charge on the cell capacitor.

@@

To read a row, the word line connected to the row's transistors must first be activated. When the word line is activated, the transistors connect the capacitors to the bit lines. When the capacitors connect to the bit lines, charge will either flow from or to the capacitors depending on whether they were charged or not, respectively. For this to work, the bit lines have to be precharged to $V_{DD}/2$. The sense amplifiers then detect the change of voltage, which makes it possible to interpret the stored data. The charge stored on the capacitors has, as mentioned, left the capacitor and a read operation is thus destructive. To maintain the data the capacitors are recharged by the bit lines, which are driven to either $V_{DD}/2$ or $0\:V$ by the sense amplifiers depending on the former voltage detected. Finally, when another row on the bank is accessed, the open row has to be closed, which disconnects the capacitors from the bit lines.

To write to a row, the corresponding word line is first activated. As during a read, the transistors then connect the capacitors to the bit lines, and charge will either flow from or to the capacitor. The row buffer is then loaded with the data that are driven to the bit lines by the sense amplifiers, thus storing the information on the capacitors. The row is then closed when another row is accessed.

@@

To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the $64\:ms$ interval. This approach is called \textit{burst refresh} and blocks normal accesses during the burst time, where power consumption peaks. A better alternative is to use \textit{distributed refresh}, which scatter the refreshes across the interval, resulting in a lower average latency for the processor and more stable power consumption.

The implementation of a DRAM refresh cycle can be made in two different ways, to either affect one row or many. The first is called \textit{RAS-only refresh} (ROR) where a refresh cycle refreshes one targeted row. To target one row, the row's address is put on the bus, and the MC has the responsibility that a row is refreshed within the maximum refresh period. The second is called \textit{CAS before RAS refresh} (CBR). In this scheme, the memory module has an internal address counter that hold which row that will be refreshed. On a refresh cycle, the row stated by the internal address counter is refreshed and the counter is then incremented. Contrary to ROR, CBR do not need provided addresses on the bus, which results in a lower power consumption and less bus usage. % During CBR, a row in each bank is updated, right? if so. write that in the text.

### 2014-10-16 paper/include/introduction.tex

Energy consumption in modern computer system is an increasingly big system design constraint. For example, battery powered mobile systems are getting more common and they needs to be energy efficient to have a long battery life. Moreover, in large data centers the power consumption corresponds to a large part of the upkeep cost.

DRAM is the most commonly used main memory technology in modern computer systems and represents a growing portion of system power consumption \cite{exascale}. This portion will continue to increase as we strive for higher and higher memory capacity. A large part of the DRAM power consumption comes from refreshing all memory-cells in the DRAM. This refreshing has to be done as DRAM's store data as a charge in a capacitor, and these charges leak over time. To increase DRAM performance and lower access latencies, 3D stacked DRAM has been proposed, but stacked design have trouble dissipating heat. The higher temperature increase leakage, and the DRAM will have to be refreshed more often.

>>>>

Energy consumption in modern computer system is an increasingly big system design constraint. For example, battery powered mobile systems are getting more common and they need to be energy efficient to have a long battery life. Moreover, in large data centers the power consumption corresponds to a large part of the upkeep cost.

DRAM is the most commonly used main memory technology in modern computer systems and represent a growing portion of system power consumption \cite{exascale}. This portion will continue to increase as we strive for higher and higher memory capacity. A large part of the DRAM power consumption originates from refreshing all DRAM memory-cells. This refreshing has to be done as DRAM's store data as a charge in a capacitor, and these charges leak over time. To increase DRAM performance and lower access latencies, 3D stacked DRAM has been proposed, but stacked design have trouble dissipating heat. The higher temperature increases leakage, and the DRAM will have to be refreshed more often.
